# Working with Ribbons (AUC, CI, etc.) {#ribbons}

```{r data-import, include=FALSE}
library(ggplot2)
library(tidyverse)
library(extrafont)
chic <- readr::read_csv("https://raw.githubusercontent.com/rana2hin/ggplot_guide/master/chicago_data.csv")
```

This is not a perfect dataset for demonstrating this, but using ribbon can be useful. In this example we will create a 30-day running average using the filter() function so that our ribbon is not too noisy.

```{r ribbon-default}
chic$o3run <- as.numeric(stats::filter(chic$o3, rep(1/30, 30), sides = 2))

ggplot(chic, aes(x = date, y = o3run)) +
   geom_line(color = "chocolate", lwd = .8) +
   labs(x = "Year", y = "Ozone")
```

How does it look if we fill in the area below the curve using the `geom_ribbon()` function?

```{r geom-ribbon-AUC}
ggplot(chic, aes(x = date, y = o3run)) +
   geom_ribbon(aes(ymin = 0, ymax = o3run),
               fill = "orange", alpha = .4) +
   geom_line(color = "chocolate", lwd = .8) +
   labs(x = "Year", y = "Ozone")
```

Nice to indicate the [area under the curve (AUC)](https://en.wikipedia.org/wiki/Area_under_the_curve_(pharmacokinetics)) but this is not the conventional way to use `geom_ribbon()`.

ðŸ’ *And actually a nicer way to achieve the same is `geom_area()`.*

<details>

<summary>Expand to see example.</summary>

```{r geom-area-AUC}
ggplot(chic, aes(x = date, y = o3run)) +
   geom_area(color = "chocolate", lwd = .8,
             fill = "orange", alpha = .4) +
   labs(x = "Year", y = "Ozone")
```

</details>

Instead, we draw a ribbon that gives us one standard deviation above and below our data:

```{r geom-ribbon-range}
chic$mino3 <- chic$o3run - sd(chic$o3run, na.rm = TRUE)
chic$maxo3 <- chic$o3run + sd(chic$o3run, na.rm = TRUE)

ggplot(chic, aes(x = date, y = o3run)) +
   geom_ribbon(aes(ymin = mino3, ymax = maxo3), alpha = .5,
               fill = "darkseagreen3", color = "transparent") +
   geom_line(color = "aquamarine4", lwd = .7) +
   labs(x = "Year", y = "Ozone")
```
