# The Dataset {#data}

We are utilizing data from the _National Morbidity and Mortality Air Pollution Study_ (NMMAPS), focusing specifically on data pertaining to Chicago and the years 1997 to 2000 to ensure manageability of the plots. For a more comprehensive understanding of this dataset, readers can refer to Roger Peng's book [Statistical Methods in Environmental Epidemiology with R](http://www.springer.com/de/book/9780387781662).

To import the data into our R session, we can employ `read_csv()` from the `{readr}` package. Subsequently, we'll store the data in a variable named `chic` using the _assignment arrow_ `<-`. Just Copy and Paste the following code.

```{r data-import}
chic <- readr::read_csv("https://raw.githubusercontent.com/rana2hin/ggplot_guide/master/chicago_data.csv")
```

💡 **The `::` symbolizes _namespace_ and enables accessing a function without loading the entire package. Alternatively, you could load the `readr` package first using `library(readr)` and then execute `chic <- read_csv(...)` subsequently.**

```{r data-inspection}
tibble::glimpse(chic)
```
```{r}
library(gt)
head(chic, 10) %>% gt()
```
