# The `ggplot2` Package {#ggplot}

`ggplot2` is a graphics system that facilitates the declarative creation of visualizations, founded on principles outlined in [The Grammar of Graphics](https://link.springer.com/chapter/10.1007/978-3-642-21551-3_13). With `ggplot2`, you furnish the data, specify how variables should be mapped to aesthetics, define graphical parameters to employ, and the system handles the rest.

A ggplot is made up several key elements:

1.  **Data**: Your raw dataset that you want to visualize.
2.  **Geometries** `geom_`: These are the shapes that represent your data, like points, lines, or bars.
3.  **Aesthetics** `aes()`: This controls how your data is visually represented, including aspects like color, size, and shape.
4.  **Scales** `scale_`: These map the data onto the aesthetic dimensions, like converting data values to plot dimensions or factor values to colors.
5.  **Statistical transformations** `stat_`: These are statistical summaries of your data, such as calculating quantiles or fitting curves.
6.  **Coordinate system** `coord_`: This defines how your data coordinates are mapped onto the plot's coordinate system.
7.  **Facets** `facet_`: This organizes your data into a grid of plots based on specified variables.
8.  **Visual themes** `theme()`: These set the overall appearance of your plot, covering things like background, grids, axes, default fonts, sizes, and colors.

::: {.callout-note appearance="simple"}
**The number of elements may vary depending on the situation you're working on.**
:::

## A Default ggplot {#default}

Before diving into the capabilities of `{ggplot2}`, we need to load the package. Alternatively, we can load it through the [tidyverse package collection](https://www.tidyverse.org/):

```{r library}
library(ggplot2)
# Or,
library(tidyverse)
```

```{r data-import, include=FALSE}
chic <- readr::read_csv("https://raw.githubusercontent.com/rana2hin/ggplot_guide/master/chicago_data.csv")
```

The syntax of `{ggplot2}` differs from base R. Following the basic elements, a default ggplot requires specifying three things: the _data_, _aesthetics_, and a _geometry_. To begin defining a plotting object, we call `ggplot(data = df)`, indicating that we'll work with that dataset. Typically, we aim to plot two variables---one on the x-axis and one on the y-axis. These are _positional aesthetics_, so we add `aes(x = var1, y = var2)` to the `ggplot()` call (where `aes()` denotes aesthetics). However, there are cases where one may need to specify one, three, or more variables, which we'll address later.

::: {.callout-important}
## Pay Attention!
We indicate the data _outside_ of `aes()` and include the variables that ggplot maps to the aesthetics _inside_ of `aes()`.
:::

In this instance, we assign the variable `date` to the x-position and the variable `temp` to the y-position. Subsequently, we'll also map variables to various other aesthetics such as color, size, and shape.

```{r ggplot}
(g <- ggplot(chic, aes(x = date, y = temp)))
```

Ah, the reason only a panel is generated when executing this code is because `{ggplot2}` lacks information on how we want to visualize the data. We still need to specify a geometry!

In `{ggplot2}`, you can store the current `ggobject` in a variable of your choosing, such as `g`. This allows you to extend the `ggobject` later by adding additional layers, either all at once or by assigning it to the same or another variable.

::: {.callout-tip title="A Quick Tip!"}
By using parentheses when assigning an object, the object will be printed immediately. Instead of writing `g <- ggplot(...)` followed by `g`, we can simply write `(g <- ggplot(...))`.
:::

There's a wide array of geometries in `{ggplot2}`, often referred to as _geoms_ because their function names typically start with `geom_`. You can find the full list of default geoms [here](https://ggplot2.tidyverse.org/reference/), and there are even more options available through extension packages, which you can explore [here](https://exts.ggplot2.tidyverse.org/). To instruct `{ggplot2}` on the style we want to use, we can, for example, add `geom_point()` to create a scatter plot:

```{r ggplot-default}
g + geom_point()
```

Great! However, this data could also be represented as a line plot (although it might not be the optimal choice, but it's a common practice). So, we can simply replace `geom_point()` with `geom_line()` and boom!

```{r ggplot-default-line}
g + geom_line()
```

Indeed, one can combine multiple geometric layers, and this is where the magic and fun truly begin!

```{r ggplot-default-line-point}
g + geom_line() + geom_point()
```

That's enough discussion on geometries for now. Don't worry, we'll dive into various plot types at a later point, as outlined [here](#charts).

### Change Properties of Geometries

Within the `geom_*` command, you can already manipulate visual aesthetics such as the color, shape, and size of your points. Let's transform all points into large fire-red diamonds!

```{r ggplot-default-col-size-shape}
g + geom_point(color = "firebrick", shape = "diamond", size = 2)
```

::: {.callout-note}
## Color or Colour?
`{ggplot2}` understands both `color` and `colour` as well as the short version `col`.
:::

::: {.callout-tip collapse="true"}
## Color Presets 🎨

<i>You can utilize preset colors (a full list can be found [here](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf)) or [hex color codes](https://www.techopedia.com/definition/29788/color-hex-code), both enclosed in quotes. Additionally, you can specify RGB/RGBA colors using the `rgb()` function. Click to expand:</i>

```{r ggplot-default-col-size-hex-rgb, eval=FALSE}
g + geom_point(color = "#b22222", shape = "diamond", size = 2)
g + geom_point(color = rgb(178, 34, 34, maxColorValue = 255), shape = "diamond", size = 2)
```

```{r ggplot-default-col-size-hex-rgb-plot, echo=FALSE, purl=FALSE}
library(patchwork)
a <- g + geom_point(color = "#b22222", shape = "diamond", size = 2)
b <- g + geom_point(color = rgb(178, 34, 34, maxColorValue = 255), shape = "diamond", size = 2)
a + b
```

:::

Each geom has its unique properties, referred to as _arguments_, and the same argument might produce different effects depending on the geom you're employing.

```{r ggplot-default-line_col-size-shape}
g + geom_point(color = "firebrick", shape = "diamond", size = 2) +
    geom_line(color = "firebrick", linetype = "dotted", lwd = .3)
```

### Replace the default `ggplot2` theme

To further demonstrate ggplot's versatility, let's enhance the appearance by removing the default grayish `{ggplot2}` style and setting a different built-in theme, such as `theme_bw()`. By using `theme_set()`, all subsequent plots will adopt the same black-and-white theme. This adjustment will notably enhance the appearance of the red points!

```{r remove-gray-background}
theme_set(theme_bw())

g + geom_point(color = "firebrick")
```

For further details on using built-in themes and customizing themes, refer to the section ["Working with Themes"](#themes). Starting from the next chapter, we'll also utilize the `theme()` function to customize specific elements of the theme.

::: {.callout-important}
## Remember!
`theme()` is a crucial command for manually adjusting various theme elements such as texts, rectangles, and lines.
:::

To explore the numerous details of a ggplot theme that can be modified, refer to the extensive list available [here](https://ggplot2.tidyverse.org/reference/theme.html). Take your time, as it's a comprehensive list!
