# Working with Margins {#margins}

```{r data-import, include=FALSE}
library(ggplot2)
library(tidyverse)
chic <- readr::read_csv("https://raw.githubusercontent.com/rana2hin/ggplot_guide/master/chicago_data.csv")
```

Sometimes it is useful to add a little space to the plot margin. Similar to the previous examples, we can use an argument to the `theme()` function. In this case, the argument is `plot.margin`. As illustrated in the previous example where we changed the background color using `plot.background`, we can now add extra space to both the left and right.

The `plot.margin` argument can handle a variety of different units (cm, inches, etc.), but it requires the use of the `unit` function from the package `grid` to specify the units. You can either provide the same value for all sides (easiest via `rep(x, 4)`) or particular distances for each. Here, I am using a 1cm margin on the top and bottom, 3 cm margin on the right, and an 8 cm margin on the left.

```{r margin}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = "Temperature (Â°F)") +
  theme(plot.background = element_rect(fill = "gray60"),
        plot.margin = margin(t = 1, r = 3, b = 1, l = 8, unit = "cm"))
```

ðŸ’ <i>You can also use `unit()` instead of `margin()`. Expand to see example.</i>

```{r margin-2}
ggplot(chic, aes(x = date, y = temp)) +
  geom_point(color = "firebrick") +
  labs(x = "Year", y = "Temperature (Â°F)") +
  theme(plot.background = element_rect(fill = "gray60"),
        plot.margin = unit(c(1, 3, 1, 8), "cm"))
```
